Red Wine Quality by Ahmad Almutawa
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load all of the packages that you end up using in your analysis in this code
# chunk.



# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library("dplyr")
library(pander)
library(ggthemes)
library(gridExtra)
library(corrplot)
library(tidyr)
```

```{r echo=FALSE,}

setwd("C:/Users/Emanice/Desktop/UDACITY - DAND/Project 5 R/Project 5")

rwine <- read.csv("wineQualityReds.csv")

# Load the Data

```

This tidy data set contains 1,599 red wines with 11 variables on the chemical 
properties of the wine. At least 3 wine experts rated the quality of each wine, 
providing a rating between 0 (very bad) and 10 (very excellent).
The data set contains 1599 number of observations with 13 variables as follows:

1 - fixed acidity: most acids involved with wine or fixed or nonvolatile (do 
not evaporate readily)

2 - volatile acidity: the amount of acetic acid in wine, which at too high of 
levels can lead to an unpleasant, vinegar taste

3 - citric acid: found in small quantities, citric acid can add 'freshness' and 
flavor to wines

4 - residual sugar: the amount of sugar remaining after fermentation stops, 
it's rare to find wines with less than 1 gram/liter and wines with greater than
45 grams/liter are considered sweet

5 - chlorides: the amount of salt in the wine

6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between
molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial
growth and the oxidation of wine

7 - total sulfur dioxide: amount of free and bound forms of S02; in low
concentrations, SO2 is mostly undetectable in wine, but at free SO2
concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

8 - density: the density of water is close to that of water depending on the
percent alcohol and sugar content

9 - pH: describes how acidic or basic a wine is on a scale from 0 (very acidic)
to 14 (very basic); most wines are between 3-4 on the pH scale

10 - sulphates: a wine additive which can contribute to sulfur dioxide gas (S02)
levels, wich acts as an antimicrobial and antioxidant

11 - alcohol: the percent alcohol content of the wine Output variable (based on
sensory data)

12 - quality (score between 0 and 10)

Please note that the first column (x) is just the rows numbers

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Exploring the data set
str(rwine)
summary(rwine)



rwine$X = factor(rwine$X)
# Changing the variable for row numbers to factor

theme_set(theme_minimal(24))
# Setting the theme for better graphing

summary(rwine$fixed.acidity)

p1 <- ggplot(aes(x = fixed.acidity), data = rwine) +
  geom_histogram(binwidth = .1) +
  scale_x_continuous(breaks = seq(4 , 20 , 1)) +
  scale_y_continuous(breaks = seq(0 , 100 , 5))

p1

```  

The graph seems to start from 4 to 16 where the mean and the median are both
between 7 to 9. This seems like a normal distribution but a little bit skewed
to the left side

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(rwine$volatile.acidity)

p2 <- ggplot(aes(x = volatile.acidity), data = rwine) +
  geom_histogram(binwidth = .01) +
  scale_x_continuous(breaks = seq(0 , 2 , .1)) +
  scale_y_continuous(breaks = seq(0 , 75 , 10))

p2

```  

The graph seems to start from 0 to 2 where the mean and the median are both
between 0.5 to 0.6 . There are some clear outliers after 1.15 till 1.6

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(rwine$citric.acid)

p3 <- ggplot(aes(x = citric.acid), data = rwine) +
    geom_histogram(binwidth = .01) +
    scale_x_continuous(breaks = seq(0 , 1 , .1)) +
    scale_y_continuous(breaks = seq(0 , 200 , 10))

p3
 
```
 
This graph seems to start from 0 to 1 where the mean and the median are both
between 0.25 to 0.3 . There is one outlier at 1

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(rwine$residual.sugar)

p4 <- ggplot(aes(x = residual.sugar), data = rwine) +
  geom_histogram(binwidth = .1) +
  scale_x_continuous(breaks = seq(0 , 20 , 1)) +
  scale_y_continuous(breaks = seq(0 , 200 , 10))

p4

```

This graph seems to start from 0 to 16. It's clearly left skewered and
outliers starts to show after 10.5

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(rwine$chlorides)
 
p5 <- ggplot(aes(x = chlorides), data = rwine) +
  geom_histogram(binwidth = .005) +
  scale_x_continuous(breaks = seq(0 , 1 , .1)) +
  scale_y_continuous(breaks = seq(0 , 300 , 25))

p5

```

This graph seems to start from 0 to 0.7 . It's clearly left skewered as the
3rd quartile is at 0.09 . 

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(rwine$free.sulfur.dioxide)
 
p6 <- ggplot(aes(x = free.sulfur.dioxide), data = rwine) +
  geom_histogram(binwidth = .5) +
  scale_x_continuous(breaks = seq(0 , 100 , 5)) +
  scale_y_continuous(breaks = seq(0 , 150 , 10))



p6

```

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

sort(rwine$free.sulfur.dioxide)

```
 
This graph is also left skewered with outliers after 65 . it's worth to
mention that the graph shows some between bars as it seems there is only one
decimal value (5.5) in this variable from 0 until the 3rd quartile

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(rwine$total.sulfur.dioxide)
 
p7 <- ggplot(aes(x = total.sulfur.dioxide), data = rwine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(0 , 300 , 25)) +
  scale_y_continuous(breaks = seq(0 , 50 , 5))

p7

```

Another left skewered with 2 outliers from  275 to 300 . The median and mean
are located from 35 to 50

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(rwine$density)

p8 <- ggplot(aes(x = density), data = rwine) +
  geom_histogram(binwidth = .0001) +
  scale_x_continuous(breaks = seq(0 , 2 , .1)) +
  scale_y_continuous(breaks = seq(0 , 50 , 5))

p8

```

This graph seems to be normally distributed and all values are located
between 0.9901 and 1.0037

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(rwine$pH)

p9 <- ggplot(aes(x = pH), data = rwine) +
  geom_histogram(binwidth = .01) +
  scale_x_continuous(breaks = seq(0 , 5 , .1)) +
  scale_y_continuous(breaks = seq(0 , 75 , 5))

p9

```
 
This graph seems to be also normally distributed especially if we exclude the
outliers

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}

summary(rwine$sulphates)

p10 <- ggplot(aes(x = sulphates), data = rwine) +
  geom_histogram(binwidth = .01) +
  scale_x_continuous(breaks = seq(0 , 5 , .2)) +
  scale_y_continuous(breaks = seq(0 , 75 , 5))

p10

```
  
A left skewered graph and clear outliers from 1.4 to 2.0 .

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(rwine$alcohol)

p11 <- ggplot(aes(x = alcohol), data = rwine) +
  geom_histogram(binwidth = .075) +
  scale_x_continuous(breaks = seq(0 , 20 , 1)) +
  scale_y_continuous(breaks = seq(0 , 200 , 10))

p11

```

The graph seems to start from 8.4 to 14.9 where the mean and the median are
both between 10 and 10.5 . This seems like a bit of a left skewered
distribution. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(rwine$quality)
 
p12 <- ggplot(aes(x = quality), data = rwine) +
   geom_histogram(binwidth = .5, fill = '#5760AB') +
   scale_x_continuous(breaks = seq(0 , 10 , 1)) +
   scale_y_continuous(breaks = seq(0 , 750 , 50))

p12

```

Although the rating options are from 1 to 10. However all the observations of
this variable is located in 3-8 where the mean and median are in the middle
5-6

```{r echo=FALSE, message=FALSE, warning=FALSE}

grid.arrange(p1 , p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, ncol =3)

```

Showing the 12 graphs altogether

```{r echo=FALSE, message=FALSE, warning=FALSE}

rwine$quality <- factor(rwine$quality, ordered = T)


rwine$rating <- ifelse(rwine$quality < 5, 'bad', ifelse(
  rwine$quality < 7, 'average', 'good'))

rwine$rating <- ordered(rwine$rating,
                       levels = c('bad', 'average', 'good'))


ggplot(aes(x = rating), data = rwine) +
  stat_count(width = 1, color = 'black',fill = I('blue'))

```
  
The graph shows that most rating for wine quality are average


# Univariate Analysis

### What is the structure of your dataset?
1,599 red wines with 12 variables and I added extra variable which is Rating and this variable is based on  the variable quality

### What is/are the main feature(s) of interest in your dataset?
Quality and rating

### What other features in the dataset do you think will help support your \investigation into your feature(s) of interest?
pH, alcohol and fixed acidity

### Did you create any new variables from existing variables in the dataset?
yes I created "rating" variable from "quality"

### Of the features you investigated, were there any unusual distributions?
Most of the variables have outliers.

### Did you perform any operations on the data to tidy, adjust, or change the form \ of the data? If so, why did you do this?
Yes, I used breaks to make the graphs show all observations of the variables


# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}

rwinecorr <- cor(rwine %>%
    dplyr::select(-X) %>%
    dplyr::select(-rating) %>%
    mutate(quality = as.numeric(quality)))
# Since we don't want to include the variable X in the graph, we removed it as 
# well as removing rating as it's a categorical variable. Moreover, we
# transferred quality to numeric again as this method only accepts numeric 
# variables

corrplot(rwinecorr, method="number")

```

From the correaltion matrix, we can see that there are no strong relations 
between any of the variables. Moreover, there are moderate positive 
correlations between some of the variables like citric acid & fixed acidity, 
density & fixed acidity, fixed acidity & citric acid. And for the moderate 
negative correlations we have for example citric acid & volatile acidity, 
pH & fixed acidity.


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}

bp1 <- ggplot(aes(x = pH, y = alcohol), data = rwine) +
  geom_point() +
  geom_smooth(method = 'lm', color = 'red')

bp1
  


```

```{r message=FALSE, warning=FALSE, include=FALSE}

cor.test(rwine$pH, rwine$alcohol, method = 'pearson')

```

There is no meningful relation between the two variables since the 
correlation value is 0.2
 
```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
bp2 <- ggplot(aes(x = chlorides, y = alcohol), data = rwine) +
  geom_point() +
  geom_smooth(method = 'lm', color = 'red')
  
bp2

```

```{r message=FALSE, warning=FALSE, include=FALSE}

cor.test(rwine$chlorides, rwine$alcohol, method = 'pearson')

```

There is no meningful relation between the two variables since the 
correlation value is -0.2

```{r echo=FALSE, message=FALSE, warning=FALSE}

bp3 <- ggplot(aes(x = chlorides, y = pH), data = rwine) +
  geom_point() +
  geom_smooth(method = 'lm', color = 'red')
  
bp3



```

```{r message=FALSE, warning=FALSE, include=FALSE}

cor.test(rwine$chlorides, rwine$pH, method = 'pearson')

```

There is no meningful relation between the two variables since the 
correlation value is -0.2

```{r echo=FALSE, message=FALSE, warning=FALSE}

bp4 <- ggplot(aes(x = density, y = pH), data = rwine) +
  geom_point() +
  geom_smooth(method = 'lm', color = 'red')
  
bp4

```

```{r message=FALSE, warning=FALSE, include=FALSE}

cor.test(rwine$density, rwine$pH, method = 'pearson')

```

There's a small negative  correlation between the two variables

```{r message=FALSE, warning=FALSE}

bp5 <- ggplot(aes(x = density, y = alcohol), data = rwine) +
  geom_point(alpha = .5, position = position_jitter(h = 0)) +
  geom_smooth(method = 'lm', color = 'red')

bp5

```

```{r message=FALSE, warning=FALSE, include=FALSE}

cor.test(rwine$density, rwine$alcohol, method = 'pearson')

```

There's a moderate negative  correlation between the two variables

```{r echo=FALSE, message=FALSE, warning=FALSE}

bp6 <- ggplot(data = rwine, aes(x = quality, y = fixed.acidity)) +
  geom_boxplot()
  
bp6


```

```{r message=FALSE, warning=FALSE, include=FALSE}

cor.test(as.numeric(rwine$quality), rwine$fixed.acidity, method = 'pearson')

```

The median values are close to each others except for 7 which comes higher 
than the rest

```{r}

bp7 <- ggplot(data=rwine, aes(x=quality, y=alcohol)) +
  geom_boxplot()

bp7

```

the more quality goes up, the more alcohol increase

```{r}

grid.arrange(bp1 , bp2, bp3, bp4, bp5, bp6, bp7, ncol =3)

```

# Showing the 7 graphs altogether



# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \ investigation. How did the feature(s) of interest vary with other features in \the dataset?
Some of them have moderate correlations while others have no correlations. Also,
there are no strong correlations found between any of the varaibles

### Did you observe any interesting relationships between the other features \ (not the main feature(s) of interest)?
I notice that there's a moderate negative correlation between citric acid &
volatile acidity

### What was the strongest relationship you found?
the strongest is between pH and fixed acidity (moderate negative with -0.68
correlation factor)

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots, message=FALSE, warning=FALSE}

mv1 <- ggplot(data = rwine,
       aes(y = fixed.acidity, x = pH,
           color = quality)) +
  geom_point() +
  scale_color_brewer()

mv1
  
fmv1 <- ggplot(aes(y = fixed.acidity, x = pH), data = subset(rwine, quality <=5)) +
  geom_line(aes(color = quality), stat= 'summary', fun.y=median)
  
fmv2 <- ggplot(aes(y = fixed.acidity, x = pH), data = subset(rwine, quality >5)) +
  geom_line(aes(color = quality), stat= 'summary', fun.y=median)

grid.arrange(fmv1, fmv2, ncol =2)

```

Alot of flactuations and intersections between different wine qualities

```{r echo=FALSE, message=FALSE, warning=FALSE}

mv2 <- ggplot(data = rwine,
       aes(y = fixed.acidity, x = citric.acid,
           color = quality)) +
  geom_point() +
  scale_color_brewer()
  
mv2

cmv1 <- ggplot(aes(y = fixed.acidity, x = citric.acid), 
               data = subset(rwine, quality <=5)) +
  geom_line(aes(color = quality), stat= 'summary', fun.y=median)
  
cmv2 <- ggplot(aes(y = fixed.acidity, x = citric.acid), 
               data = subset(rwine, quality >5)) +
  geom_line(aes(color = quality), stat= 'summary', fun.y=median)
  
grid.arrange(cmv1, cmv2, ncol =2)

```
Alot of flactuations except for wine with the lowest quality

```{r echo=FALSE, message=FALSE, warning=FALSE}

mv3 <- ggplot(data = rwine, aes(y = fixed.acidity,
                                x = density, color = quality)) +
  geom_point() +
  scale_color_brewer()

mv3

```

alternative plot of the same variables

```{r echo=FALSE, message=FALSE, warning=FALSE}

amv1 <- ggplot(aes(y = fixed.acidity, x = density),
               data = subset(rwine, quality <=8)) +
  geom_point(aes(color = quality), stat= 'summary', fun.y=median)

amv1

```

seems that red wine with high fixed acidity tend to have better quality than
most of the red wine with low quality

```{r echo=FALSE, message=FALSE, warning=FALSE}

mv4 <- ggplot(data = rwine, aes(y = pH, x = alcohol, color = quality)) +
  geom_point() +
  scale_color_brewer()

mv4

```
  
There's no clear correlation in this graph

```{r echo=FALSE, message=FALSE, warning=FALSE}

mv5 <- ggplot(data = rwine, aes(y = chlorides, x = alcohol, color = quality)) +
  geom_point() +
  scale_color_brewer()

mv5

```

There's no clear correlation in this graph 

```{r echo=FALSE, message=FALSE, warning=FALSE}

mv6 <- ggplot(data = rwine, aes(y = density, x = alcohol, color = quality)) +
  geom_point() +
  scale_color_brewer()

mv6

```

There's no clear correlation in this graph 


```{r echo=FALSE, message=FALSE, warning=FALSE}

mv7 <- ggplot(data = rwine, aes(y = density, x = pH, color = quality)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating) +
  scale_color_brewer()

mv7
  
ggplot(aes(y = fixed.acidity, x = density), data = subset(rwine, quality <=8)) +
  geom_smooth(aes(color = quality))

```

There's no clear correlation in this graph 

```{r echo=FALSE, message=FALSE, warning=FALSE}

grid.arrange(mv1 , mv2, mv3, mv4, mv5, mv6, mv7, ncol =2)

```




# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \ investigation. Were there features that strengthened each other in terms of \ looking at your feature(s) of interest?
I used quality as the 3rd variable for the plots and in most of the graphs, 
I can say that quality doesn't have a clear corelation

### Were there any interesting or surprising interactions between features?
No, but I just notice alot of flactuations for some of the plots


------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

ggplot(data = rwine, aes(y = fixed.acidity, x = density, color = quality)) +
  geom_point() +
  scale_color_brewer() +
  ggtitle("Fixed acidity per density coloured by quality")

```

### Description One: 
As there is a positive correlation between fixed acidity and density, we can
also notice that the dots becomes darker as the fixed acidity increases which
indicates the increase in the quality of the wine


### Plot Two
```{r echo=FALSE, Plot_Two}

ggplot(data = rwine,
       aes(y = fixed.acidity, x = citric.acid,
           color = quality)) +
  geom_point() +
  scale_color_brewer() +
  ggtitle("Fixed acidity per citric acid coloured by quality")

```

### Description Two: 
As there is a moderate positive correlation between fixed acidity and citric
acid, we can also notice that the dots becomes darker as the citric acid
increases which indicates the increase in the quality of the wine


### Plot Three
```{r echo=FALSE}

ggplot(data=rwine, aes(x=quality, y=alcohol)) +
  geom_boxplot() +
  facet_wrap(~rating) +
  ggtitle("Alcohol concentration per quality")

```

### Description Three: 
For most of the times, Alcohol concentration increases along with higher
quality of the wine

------

# Reflection

### What were some of the struggles that you went through?

Relations were difficult to figure out

### What went well? What was surprising? 

Creating the rating variable was easy and its relation was clear with some of
the variables like with alcohol
